\documentclass[10pt,a4paper,twocolumn]{article}
\usepackage[cm]{fullpage}
\usepackage{newtxtext}
\usepackage{newtxmath}
\usepackage{listings}
\usepackage{listings-rust}
\usepackage{xcolor}

\lstset{
	commentstyle=\color{olive},
	keywordstyle=\color{blue},
	basicstyle=\footnotesize\selectfont\ttfamily,
	tabsize=2,
	captionpos=b,
	language=Rust
}

\begin{document}

\title{Advanced Systems Programming - Summarise Exercise 1}
\author{Stuart Reilly - 2258082}
\date{\today}

\maketitle

\section{Rust Memory Management}
\label{rustMem}
Rust makes use of region-based memory management rather than reference counting or a garbage
collector.
When an object is allocated, it's assigned to a region.
Once this region is deallocated, all objects assigned to the region are also deallocated.
This is similar to stack-allocations, as all objects allocated on the stack are deallocated
when the stack frame is deallocated.
An object can be assigned to a different region after it is allocated.

This is represented in Rust as lifetimes, where a region is often the scope of the current
function.
A region is deallocated when the current function terminates, so all objects allocated during
the function go out of scope and are deallocated.
The Rust compiler tracks when an object is created, when a reference is created to the object,
when a reference goes out of scope, when the object changes ownership and when the object goes
out of scope.
In Listing \ref{lst:simpleLife}, the variable \lstinline{x} has ownership of the heap-allocated
integer 42.
The function \lstinline{print_consume} takes ownership of \lstinline{x} as its parameter
\lstinline{b}, so \lstinline{b} is deallocated when the function returns, deallocating the integer.
Since the compiler tracks ownership of \lstinline{x}, the program will not compile if \lstinline{x}
is used after \lstinline{print_consume} returns.
If \lstinline{print_consume} returned \lstinline{b} and the \lstinline{main} function assigned the
return value of \lstinline{print_consume} to \lstinline{x}, then \lstinline{x} could be used after
the function returned, as ownership was returned.

\begin{lstlisting}[float=h, label=lst:simpleLife, caption={A short example of
	when a object is allocated and deallocated in Rust.}]
fn print_consume(b: Box<usize>) {
	println!("{}", b);
} // b goes out of scope and is deallocated

fn main() {
	// Allocate space on the heap for the integer 42
	let x = Box::new(42); 
	// Print x
	print_consume(x);
	// print_consume took ownership of x so it's
	// been deallocated

	// println!("{}", x); would fail to compile
}
\end{lstlisting}

Rust has three rules for ownership: each value has a variable which is its owner, each value
can have exactly one owner, and when the owner goes out of scope the value is dropped.
The first and third rules are explained above and demonstrated in Listing \ref{lst:simpleLife}.
Listing \ref{lst:owner} shows how exactly one variable can own an object.
The variable \lstinline{x}'s stack data is copied bitwise to the variable \lstinline{y}, therefore
the heap allocated integer is not copied, rather the pointer (Box) to it.
Once the stack data is copied, \lstinline{x} is no longer deemed valid, so the \lstinline{drop}
function for \lstinline{x} is not called when the function returns, avoiding a double-free.
If the compiler did not track the ownership of the object, both \lstinline{x} and \lstinline{y}
would both own the object, so the compiler forbids \lstinline{x} from being used after having the
ownership of the object being passed to \lstinline{y}.
Furthermore, if both variables where
This is known as move semantics.
Rust supports copy semantics, which is also shown in Listing \ref{lst:owner} with the variables
\lstinline{a} and \lstinline{b}.
Since the integer is not heap allocated, but stack allocated, and integers implement the
\lstinline{Copy} trait, both \lstinline{a} and \lstinline{b} remain valid until the end of the
function.
Both \lstinline{a} and \lstinline{b} have their own copy of the integer.
To facilitate creating copies of an object with move semantics, there exists the \lstinline{Clone}
trait.
A type which implements the \lstinline{Clone} trait is a type which can be deep copied via an
explicit call to the \lstinline{clone} function, provided by the \lstinline{Clone} trait.
The implementation of the \lstinline{clone} function can contain arbitrary code and as such is
assumed to be an expensive operation.
The \lstinline{Copy} trait on the other hand does not provide any methods, and is used as a marker.
A type which implements the \lstinline{Copy} trait can be copied with a simple bitwise copy of the
stack allocation for the type, and is given copy semantics.
Since this is very cheap computationally, a bitwise copy is inflict.
In Listing \ref{lst:owner}, the heap allocated integer \lstinline{x} does not implement
\lstinline{Copy}, since a bitwise copy of the stack allocation would break the ownership rules, and
lead to to variables owning the same heap allocation.
Whereas the stack allocated integer \lstinline{a} implements \lstinline{Copy} since a simple
bitwise copy of the stack allocation does not break the ownership rules.
A type must manually implement the \lstinline{Clone} and \lstinline{Copy} traits.

\begin{lstlisting}[language=Rust, float=h, label=lst:owner, caption={A short example of how there
	can be only one owner of a value in Rust.}]
fn main() {
	// Allocate space on the heap for the integer 42
	let x = Box::new(42);
	// Transfer ownership of the integer to variable y
	let y = x;
	// x cannot be used after this point
	println!("{}", y);
	// println!("{}", x); would fail to compile

	// Stack allocated integer
	let a = 64;
	// a is copied to b, so a can be used after this point
	let b = a;
	println!("{}", a);
	println!("{}", b);
} // y, a, b are deallocated
\end{lstlisting}

To allow a variable to be borrowed, and not transfer ownership, Rust has four pointer types.
Two references which follow the borrowing rules, and two raw pointers which are equivalent to C
pointers.
The borrowing rules are all references must be valid, and at any given time either one mutable
reference or any number of immutable references can exist.
The references are \lstinline{&T} and \lstinline{&mut T}, which are immutable and immutable
respectively.
As shown in Listing \ref{lst:refLife}, a reference cannot outlive the object it references and
either multiple immutable references or a single mutable reference can exist at once.
Te compiler ensures references follow the borrowing rules by ensuring references cannot be created
to an object before it is initialised, and does not live longer than the object.
As such, it is impossible to have a dangling reference or use-after-free bug.
Furthermore, the compiler keeps track of how many mutable and immutable references exist for an
object at once.
Therefore, the compiler can ensure either multiple immutable references or a single mutable
reference can exist at once, so both data races and iterator invalidation can not occur.
Both data races and iterator invalidation require either multiple mutable references or both
mutable and immutable references to exists at once.
Some data structures can not be represented with Rust's ownership rules.
For example a doubly linked list requires each node to have references to the next and previous
nodes.
If a new node was to be added to the list, the nodes either side of where the new node will go
will need to be modified, but this requires both mutable and immutable references to these nodes
to exist at once.

\begin{lstlisting}[language=Rust, float=h, label=lst:refLife, caption={A short example of the
	lifetime of references and how only immutable references or a mutable reference can exist
	at once}]
fn print(y: &Box<usize>) {
	println!("{}", y);
} // Ref y dealloacted here, not heap allocation

fn increment(z: &mut Box<usize>) {
	*z += 1;
} // Mut ref z deallocated here, not heap allocation

// This function would fail to compile
// fn generate() -> &Box<usize> {
//	let i = Box::new(64);
//	&i
// } // i deallocated here so would return invalid ref

fn main() {
	// Allocate space on the heap for the integer 42
	// Must be mut to allow mutable references
	let mut x = Box::new(42);

	increment(&mut x);
	print(&x);
	// x still exists here

	{
		let a = &x;
		let b = &x;
		// Above is valid because both immutable
		// let bad = &mut x; would fail to compile
	} // a and b deallocated, not x

	{
		let c = &mut x;
		// c can be the only reference of x

		// let d = &mut x;
		// let e = &x;
		// Both above would fail to compile
	} // c deallocated, not x
} // x is deallocated here
\end{lstlisting}

The raw pointers are \lstinline{*mut T} and \lstinline{*const T}, which are equivalent to C's
\lstinline{T*} and \lstinline{const T*} pointers respectively.
These do not follow the borrowing rules, so there can be an arbitrary number of \lstinline{*mut T}
and \lstinline{*T} to the same object at once.
Furthermore, they do not have any automatic cleanup, there is no guarantee they point to valid
memory and not null.
As such raw pointers can only be dereferenced by unsafe code.
They are primarily use for C FFI, and implementing certain data structures like doubly linked lists.

\section{Comparison between Rust's and C's memory management}
\label{cComp}

C requires manual memory management through the use of \lstinline{malloc} and \lstinline{free}.
This requires the programmer to keep track of the lifetime of objects.
If the programmer forgets to \lstinline{free} an object, the memory will be leaked.
Rust prevents this by ensuring all allocated memory is deallocated when it goes out of scope.
Even if an object is freed, the programmer could still introduce a double-free or use-after-free
bug.
By ensuring there are no references to an object when it is deallocated, Rust can ensure these bugs
are impossible.
Since C does not have any compiler-enforced ownership rules, both \lstinline{T*} and
\lstinline{*const T} can exist simultaneously, leading to data races and iterator invalidation.
It is impossible to have both \lstinline{&T} and \lstinline{&mut T} simultaneously in Rust,
therefore both data races and iterator invalidation are impossible in Rust.
In C the programmer has to enforce these conditions at runtime, whereas the Rust compiler enforces
them at compile time.
Therefore, a safe Rust program is one which can compile, whereas a safe C program requires a large
amount of effort to prove and enforce at runtime.
Listing \ref{lst:rustSafe} shows a small Rust program, and Listing \ref{lst:cSafe} shows the same
program in C with the imports omitted.
As shown, there is a noticeable difference between the amount of code to ensure the program is
safe, and the lack of a \lstinline{drop} method in the Rust version.

\begin{lstlisting}[language=Rust, float=h, label=lst:rustSafe, caption={A short safe Rust program
	to allocate, increment, print and deallocate an integer.}]
fn allocate() -> Box<usize> {
	Box::new(42);
}

fn print(i: &usize) {
	println!("{}", *i);
}

fn main() {
	let i = allocate();
	*i += 1;
	print(i);
}
\end{lstlisting}

\begin{lstlisting}[language=C, float=h, label=lst:cSafe, caption={A short safe C program to
	to allocate, increment, print and deallocate an integer.}]
int *allocate() {
	int *i = malloc(sizeof(int));
	if (i == NULL) {
		return NULL;
	}
	*i = 42;
	return i;
}

void print(int *i) {
	if (i == NULL) {
		return;
	}
	printf("%d\n", *i);
}

void drop(int *i) {
	free(i);
}

int main() {
	int *i = allocate();
	*i += 1;
	print(i);
	drop(i);
}
\end{lstlisting}

In terms of performance, Rust's memory management has no runtime cost over C's manual memory
management.
The Rust compiler keeps track of when objects should be freed, and inserts the code to free the
object at that point.
With that said, there can be argued a runtime cost to C's manual memory management over Rust's
due to the unsafe nature of C's pointers.
When a pointer is given to a function in C, it requires a null check to avoid segmentation faults,
whereas Rust's references are guaranteed to never be null, so can avoid this check.
Furthermore, once a C pointer has been given to \lstinline{free}, the programmer must either set
all pointers to the object behind the pointer to null, or manually ensure no other pointers exist.
C's \lstinline{free} can accept null pointers, so changing a freed pointer to null prevents
double-free bugs.
Since the Rust compiler keeps track of when an object is freed and ensures no references exist,
both the runtime cost of changing pointers to null and manually tracking pointers can be avoided.

Due to Rust's borrowing rules, some data structures can not be represented in safe rust, for
example doubly linked lists as discussed in Section \ref{rustMem}.
These data structures must be implemented in unsafe code and utilise Rust's raw pointers.
Whereas in C, there are no borrowing rules, so these data structures can be implemented without
any extra ceremony.

\section{Comparison between Region-based and Garbage-collected memory management}

Garbage-collected memory management is where the runtime keeps track of the references between all
objects.
Periodically, the program is interrupted while a garbage collector removes all objects which are no
longer referenced.
Therefore, garbage-collected languages push the memory management to a runtime cost, unlike Rust
which forces the programmer to handle memory management at compile time.
So, writing a program in a garbage collected language is easier than in Rust, in general.

The more optimal algorithms for removing garbage objects require double the memory usage.
These algorithms allocated two regions, one which is currently being used for allocations, and
another which will be used for allocations after the next garbage collection.
During garbage collection all objects, which still have live references to it, are moved to the
unused region, then the references to these objects are updated, and the first region is
deallocated.
Since two regions need to be allocated for these algorithms, overall memory usage is much higher
than Rust's region-based memory management.

As discussed in Section \ref{cComp}, safe Rust can not represent a doubly linked list.
Since garbage-collected languages do not have mutability borrowing rules like Rust, so they can
implement a safe doubly linked list.
Listing \ref{java} shows an example of a queue implemented in Java as a doubly linked list.
In order to do this in Rust, \lstinline{unsafe} must be used with raw pointers.

\begin{lstlisting}[language=Java, caption={An example of a queue in Java implemented as a doubly
	linked list.}, label=java]
public class LinkedList<T> {
	private Node<T> head;
	private Node<T> tail;

	public LinkedList<T>(T first) {
		Node<T> node = new Node<T>(first, null, null);
		head = node;
		tail = node;
	}

	public void push(T data) {
		if (tail == null) {
			tail = new Node<T>(data, null, null);
			tail = head;
		} else {
			Node<T> oldTail = tail;
			tail = new Node<T>(data, null, oldTail);
			oldTail.next = tail;
		}
	}

	public T pop() {
		if (head == null) {
			return null;
		}
		Node<T> oldHead = head;
		head = head.next;
		head.prev = null;
		return head.data;

	}

	private class Node<T> {
		protected T data;
		protected Node<T> next;
		protected Node<T> prev;

		protected Node<T>(T data, Node<T> next, Node<T> prev) {
			this.data = data;
			this.next = next;
			this.prev = prev;
		}
	}
}
\end{lstlisting}

Another issue with garbage collection is they do not work efficiently with caches and virtual
memory.
When the garbage collector is called, the runtime performs a context switch to the garbage
collector.
This will invalidate all CPU caches by the time the program is switched back to.
Furthermore, the garbage collector is likely to require access to pages which the operating
system could have swapped to disk while the program was running.
This will lead to page faults when switching to and from the garbage collector.

Since there is no guarantee when the garbage collector is called, the amount of actively used
by the program is non-deterministic.
Whereas Rust can determine when to release memory at compile time, the memory usage of the program
is deterministic.

\end{document}
